---
    title: "RunATAC"
    author: "Sam Buckberry"
    date: "`r format(Sys.Date())`"
    output: github_document
---

RunATAC
---
## An R package for processing and plotting ATAC-seq data

This package is currently is it's infancy. Don't expect any real functionality for a while...

You can use RunATAC to easily
* Plot insert size histograms 
* Calculate fraction of reads in peaks (FrIP)
* Plot footprint profiles for motifs 
* Generate V-plots of fragment sizes flanking transcription factor motifs
* Generate bigwig files of Tn5 insertions and nucleosome spanning fragments
* Perform motif-centric transcription factor occupancy predictions with the CENTIPEDE algorithm
* Generate tables of counts for multiple samples
* Perform differential accessibility testing
* Generate browser snapshots for regions of interest

###TODO
* Browser shots
* Peak calling with MACS2 using generated bigwigs
* Nucleosome density plot analogous to the footprint plot
* Footprint and nucleosome together
* Heat maps similar to deeptools


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, results = "hide",
                      comment = FALSE, warning = FALSE, message=FALSE)
```

### Install the RunATAC package and dependencies
```{r, eval=FALSE}
library(devtools)
devtools::install_github("SamBuckberry/RunATAC")

install.packages('data.table')
install.packages('magrittr')

## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite(c("Biostrings", "motifRG", "GenomicRanges", "IRanges", "GenomicAlignments", "rtracklayer", "Rsamtools"))
```

### Plot ATAC-seq insert size histogram
One of the first quality control steps of an ATAC-seq experiment often performed is the assessment of insert size distribution. As Tn5 DNA insertions in accessible chromatin are highly structured around nucleosomes [Buenrostro et. al. 2013](http://www.nature.com/nmeth/journal/v10/n12/full/nmeth.2688.html), a good ATAC-seq experiment should show an enrichment of fragment widths (the mapping distance between 5' read ends for paired-end reads) below 100 bases, and another peak of enrichment at ~200 bases, indicative of nucleosome spanning read fragments.

To inspect the insert sizes from aligned ATAC-seq reads, first load the fragment data into into a GRanges object using the `read_atac_frags()` function. Here we will use a BAM file included in the RunATAC package. This BAM is from ATAC-seq on Saccharomyces cerevisiae and includes data only from chrIV. 
```{r}
library(RunATAC)
bam_fl <- system.file("extdata", "chrIV.bam", package = "RunATAC")
frags <- read_atac_frags(bam_file = bam_fl)
```

Plot the insert size distribution. 
```{r, fig.height=3, fig.width=5}
plot_insert_size(frags, xlim=c(0,600))
```

### Calculate the fraction of Tn5 insetions in peaks (FrIP)
Another useful metric for evaluating the quality of an ATAC-seq experiment is the fraction of reads in peaks (FrIP), which can help you assess the ammount of background noise in the experiment. 

First, we load the Tn5 insertion points from BAM file into GRanges object
```{r}
ins <- read_atac_insertions(bam_fl)
```

Then we load a file of peaks in the MACS2 .narrowPeak format, however you can generate peaks using any peak caller and load these into a GRanges object from a BED format file. 
```{r}
peak_fl <- system.file("extdata", "chrIV_peaks.narrowPeak", package = "RunATAC")
peaks <- read_bed(peak_fl)
```

And now we calculate the fraction of reads in peaks
```{r}
calc_frip(signal = ins, peaks = peaks)
```

### Generate a read counts table for peaks
If the purpose of the ATAC-seq experiments is to perform differential accessibility testing, you would need to generate a table of read counts for each library.

```{r}
library(GenomicAlignments)
library(stringr)
olap_counts <- summarizeOverlaps(features = peaks, reads = ins)
olap_counts <- assays(olap_counts)$counts
rownames(olap_counts) <- str_c(seqnames(peaks), start(peaks), sep = ":") %>% 
        str_c(end(peaks), sep = "-")

```

### Calculate Tn5 insertion bias
As Tn5 has has a sequence preference [Buenrostro et. al. 2013](http://www.nature.com/nmeth/journal/v10/n12/full/nmeth.2688.html) it can be useful to calculate a Position Weight Matrix (PWM) for sequence content ±10 bp from Tn5 insertion points, so insertion counts at each position can be bias corrected. 

First, we can inspect the di-nucleotide insertion frequencies and see that there is a distinct depletion of insertions in CG dinucleotide context.
```{r, fig.width=10, fig.height=7}
library(BSgenome.Scerevisiae.UCSC.sacCer3)
dinuc_freq <- calc_dinuc_freq(ins_gr = ins, genome = Scerevisiae)
barplot(dinuc_freq$percentage, names=dinuc_freq$dinucleotide)
```

Now we construct a 21 bp positional weight matrix (PWM) for the Tn5 insertion positions and the flanking 10 bases.   
```{r}
ins_pwm <- calc_ins_pwm(ins_gr = ins, genome = Scerevisiae)
ins_pwm
```


```{r}
library(reshape2)
library(ggplot2)

dat <- melt(ins_pwm)
gg <- ggplot(dat, aes(x = Var2, y = value, group=Var1, colour=Var1)) +
        geom_line()
gg

```


### Generate a Tn5 insertion site and nucleosome bigwig files

As ATAC-seq data can provide information on both Tn5 accessible DNA and nucleosome positioning, it is useful to summarise the coverage depth of both insertions and fragment centres. For this purpose, the [bigwig file format](https://genome.ucsc.edu/goldenpath/help/bigWig.html) is very efficient for visualisation and fast data retrieval. 

To convert BAM files to insertion point and fragment centre bigwig files:
```{r, eval=FALSE}
bam_to_insertions_bw(bam_file = bam_fl, file = "inst/extdata/chrIV.ins.bigwig")
bam_to_centres_bw(bam_file = bam_fl, file = "inst/extdata/chrIV.nuc.bigwig")
```

### Plotting ATAC-seq footprints

In this example, we will plot the Tn5 insertion signal around REB1 motifs in ATAC-seq peaks using a positional weight matrix (PWM).

Get the REB1 motif from the JASPAR database.
```{r, message=FALSE, comment=FALSE}
#source("https://bioconductor.org/biocLite.R")
#biocLite(c("BSgenome.Scerevisiae.UCSC.sacCer3", "JASPAR2016", "TFBSTools"))
library(JASPAR2016)
library(TFBSTools)
getMatrixByID
tf <- getMatrixByID(JASPAR2016, ID=c("MA0363.1"))
tf <- tf@profileMatrix
```

Get the positions of motif in peaks.
```{r}
motif_pos <- motif_gr(gr = peaks, pwm = tf, genome = Scerevisiae)
```

Get the insertion data for the motif regions, for this we retreivie the data from the bigwig file of Tn5 insertions. This will return a data.frame of insertions for the 100 bases flanking each side of the motif matches.
```{r}
ins_bw <- system.file("extdata", "chrIV.ins.bigwig", package = "RunATAC")
dat <- range_summary(bigwig = ins_bw, gr = motif_pos, range = 100)
dat[1:6, 1:6]
plot(colSums(dat), type='l')
```

Then Calculate Tn5 insertion bias for the same regions
```{r}

par(mfrow=c(2,1))
plot(colSums(ins_bias), type='l')
plot(colSums(dat), type='l')
ins_over_bias <- dat+1 / ins_bias
plot(colSums(ins_over_bias))
lines(colSums(ins_bias))

ins_bw <- system.file("extdata", "chrIV.ins.bigwig", package = "RunATAC")
frag_bw <- system.file("extdata", "chrIV.nuc.bigwig", package = "RunATAC")
gr <- motif_pos
bias_pwm <- ins_pwm
genome <- Scerevisiae
range <- 200

# Get fragment centres, tn5 insertions and insertion bias scores for regions flanking PWM matches in peaks.

calc_motif_profile <- function(ins_bw, frag_bw, gr, ins_pwm, genome, range=100){
        
        # Get the range summary for Tn5 insertions
        ins_dat <- range_summary(bigwig = ins_bw, gr = gr, range = range)
        
        # Get the range summary for the fragment centres
        frag_dat <- range_summary(bigwig = frag_bw, gr = gr, range = range)
        
        # Calculate the bias scores
        bias_dat <- calc_profile_bias(ins_pwm = ins_pwm, gr = gr,
                              genome = genome, range = range)
        
        # Get data into list for export
        dat <- list(ins_dat=ins_dat, frag_dat=frag_dat, bias_dat=bias_dat)
        
        return(dat)
}


profile_dat <- calc_motif_profile(ins_bw = ins_bw, frag_bw = frag_bw, gr = motif_pos,
                                ins_pwm = ins_pwm, genome = Scerevisiae, range = 200) 


plot_motif_profiles <- function(profile_dat){
        
        range01 <- function(x){(x-min(x))/(max(x)-min(x))}
        
        ins_dat <- colSums(profile_dat[[1]]) %>% range01()
        frag_dat <- colSums(profile_dat[[2]]) %>% range01()
        bias_dat <- colSums(profile_dat[[3]]) %>% range01()
        
        plot(ins_dat)
        plot(frag_dat)
        
}


```




```{r, fig.height=4, fig.width=4.5, eval=FALSE}
gr <- motif_pos
ins_bigwig <- ins_bw
genome <- Scerevisiae

calc_footprint <- function(ins_bigwig, gr, ins_pwm, genome, flank=100){
        
        # Get the insertion data for the GRanges
        dat <- RunATAC::range_summary(bigwig = ins_bigwig, gr = gr, range = flank)

        # Calculate the insertion bias for regions
        regions_gr <- GenomicRanges::resize(gr, width = flank+(flank+1), 
                                           fix = 'center')
        
        bias <- RunATAC::calc_ins_bias(ins_pwm = ins_pwm,
                              regions_gr = regions_gr,
                              genome = genome)        
        
        bias <- do.call(rbind, bias)

        # undo the log2 transform
        bias <- exp(bias)
        
        # Calculate insertions / bias
        ins_over_bias <- dat / bias        
 
        # data for plotting
        xdat <- -flank:flank
         
        ins_y <- colSums(dat)
        bias_y <- colSums(2^bias)
        norm_y <- ins_y / bias_y

}


library(reshape2)
library(dplyr)
library(ggplot2)
mdat <- melt(dat)
fp <- mdat %>% group_by(variable) %>%
                        summarise(yy=mean(value))
fp$variable <- as.character(fp$variable) %>% as.numeric()


gg <- ggplot(fp, aes(x = variable, y = yy, fill = 'blue', title="")) +
                #stat_smooth(span = 10) +
                #geom_area(aes(fill = 'blue', stat = "bin")) +
                geom_line(size=0.3) +
                xlab("Distance from motif") +
                ylab("Tn5 insertions") +
                ggtitle(title) +
                theme_bw() +
                theme(plot.background = element_blank(),
                      panel.border = element_blank(),
                      strip.background = element_rect(size = 0),
                      panel.grid.minor = element_blank(),
                      axis.line = element_line(),
                      axis.text = element_text(size=8))
gg
```

Generate V-plot for motif regions
```{r, fig.height=4, fig.width=4.5}
v_dat <- calc_v(frags_gr = frags, motif_pos_gr = motif_pos,
                flank = 200, max_frag = 400)

plot_v(df = v_dat)
```

Predict motifs with TF occupancy using the CENTIPEDE algorithm
```{r}


```


### Plot Tn5 insertion density around transcription start sites (TSS) 
```{r}
library(GenomicFeatures)
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Scerevisiae.UCSC.sacCer3.sgdGene")
library(TxDb.Scerevisiae.UCSC.sacCer3.sgdGene)
tss <- promoters(x = TxDb.Scerevisiae.UCSC.sacCer3.sgdGene, upstream = 1, downstream = 1)
tss_profile <- range_summary(bigwig = ins_bw, gr = tss, aggregate = TRUE)
plot(tss_profile, type='l')

```

